Supermartingales and Termination Analysis of Statistical PCF

### 2021-01-16 LICS21/SPCF.tex
Introduction:ongoing Reformatted last antitone sparse ranking function in appendix

The language PPCF is a simply-typed lambda calculus with sampling of real numbers from the closed interval $[0,1]$. 
%\changed[lo]{following \cite{MakOPW21} except the score function (\Cref{rem:score})}. 
Types and terms are defined as follows, where $r$ is a real number, $x$ is a variable, $f : \mathbb{R}^n \to \mathbb{R}$ is any measurable function, and $\Gamma$ is an environment:
\begin{align*}
  \text{\em types } A, B &::= \textsf{R}  \mid  A \to B \\
  \text{\em values } V &::= \lambda x.M  \mid  \underline{r} \\
  \text{\em terms } M, N &::= V  \mid  x  \mid  M_1 \, M_2  \mid  \underline{f}(M_1,\dots ,M_n)  \mid  \tY \, M \\
   & \qquad \mid  \tif{M < 0}{N_1}{N_2}  \mid  \tsample
\end{align*}
The typing rules are standard (see \Cref{fig:typing rules}).
Terms are identified up to $\alpha$-equivalence, as usual. 
The set of all terms is denoted $\Lambda$, and the set of closed terms is denoted $\Lambda^0$.

%\paragraph{}
To define the call-by-value reduction relation, let \emph{evaluation contexts} be of the form:
\begin{align*}
  E ::= & \, [\,] \mid E \, M \mid V \, E \mid \underline{f}(\underline r_1,\dots ,\underline r_{k-1}, E, M_{k+1}, \dots, M_n) \\ & \mid \tY E \mid \tif{E<0}{M_1}{M_2}
\end{align*}
then a term reduces if it is formed by substituting a redex in a context i.e.
\begin{align*}
  E[(\lambda x.M) V] & \to E[M[V/x]] \\
  E[\underline f (\underline r_1, \dots , \underline r_n)] & \to E[\underline{f(r_1,\dots,r_n)}] \\
  E[\tY \lambda x. M] & \to E[\lambda z. M[(\tY \lambda x. M)/x] z] \\
  & \qquad \qquad \text{ where $z$ is not free in $s$}\\
  E[\tif{\underline r < 0}{M_1}{M_2}] & \to E[M_1] \text{ where }r < 0 \\
  E[\tif{\underline r < 0}{M_1}{M_2}] & \to E[M_2] \text{ where }r \geq 0 \\
  E[\tsample] & \to E[\underline r] \text{ where } r \in [0,1].
\end{align*}
We write $\to^\ast$ for the reflexive, transitive closure of $\to$.

Every closed well-typed term either is a value or reduces to another term.

@@

Define a \emph{skeleton} to be a term but, instead of having real constants $\underline r$, it has a placeholder $\skeletonPlaceholder$, so that each term $M$ has a skeleton $\mathit{Sk}(M)$, and each skeleton $S$ can be converted to a term $S[r]$ given a vector $r$ of $n$ real numbers to substitute in, where $n$ is the number of occurrences of $\skeletonPlaceholder$ in $S$. 
Following \cite{DBLP:conf/icfp/BorgstromLGS16}, the $\sigma$-algebra on $\Lambda$ is defined by identifying $\Lambda$ with $\bigcup_{m\geq 0} \big(\mathsf{Sk}_m \times \Real^m \big)$, where $\mathsf{Sk}_m$ is the set of skeletons containing $m$ occurrences of $\skeletonPlaceholder$.
Thus identified, we give $\Lambda$ the countable disjoint union topology of the product topology of the discrete topology on $\mathsf{Sk}_m$ and the standard topology on $\Real^m$, and take the corresponding Borel $\sigma$-algebra.
Note that the connected components of $\Lambda$ have the form $\{S\} \times \Real^m$, with $S$ ranging over $\mathsf{Sk}_m$, and $m$ over $\mathbb{N}$.

@@

%\paragraph{}
\begin{remark}
\label{rem:score}
Our definition of AST is equivalent to that given in \cite{MakOPW21} (although the program semantics is stated in a slightly different way), except for the presence of a $\tscore(r)$ construct for soft conditioning. 
The $\tscore(r)$ construct is irrelevant to termination except that it fails if its argument is less than or equal to zero, which allows computations to fail after finitely many steps. 
If such a thing were added, it would merely require an additional side condition to our main theorems, that no failing term was reachable (or, in the case of the confluent semantics based results, no term where the reduction strategy is not defined is reachable via the reduction strategy unless it's a value).
\end{remark}

>>>>

The language PPCF is a CbV version of PCF with sampling of real numbers from the closed interval $[0,1]$ \cite{Ehrhard2018c,DBLP:journals/pacmpl/EhrhardPT18,MakOPW21}.
%\changed[lo]{following \cite{MakOPW21} except the score function (\Cref{rem:score})}. 
Types and terms are defined as follows, where $r$ is a real number, $x$ is a variable, $f : \mathbb{R}^n \to \mathbb{R}$ is any measurable function, and $\Gamma$ is an environment:
\begin{align*}
  \text{\em types } A, B &::= \textsf{R}  \mid  A \to B \\
  \text{\em values } V &::= \lambda x.M  \mid  \underline{r} \\
  \text{\em terms } M, N &::= V  \mid  x  \mid  M_1 \, M_2  \mid  \underline{f}(M_1,\dots ,M_n)  \mid  \tY \, M \\
   & \qquad \mid  \tif{M < 0}{N_1}{N_2}  \mid  \tsample
\end{align*}
The typing rules are standard (see \Cref{fig:typing rules}).
Terms are identified up to $\alpha$-equivalence, as usual. 
The set of all terms is denoted $\Lambda$, and the set of closed terms is denoted $\Lambda^0$.

%\paragraph{}
To define the call-by-value reduction relation, let \emph{evaluation contexts} be of the form:
\begin{align*}
  E ::= & \, [\,] \mid E \, M \mid V \, E \mid \underline{f}(\underline r_1,\dots ,\underline r_{k-1}, E, M_{k+1}, \dots, M_n) \\ & \mid \tY E \mid \tif{E<0}{M_1}{M_2}
\end{align*}
then a term reduces if it is formed by substituting a redex in a context i.e.
\begin{align*}
  E[(\lambda x.M) V] & \to E[M[V/x]] \\
  E[\underline f (\underline r_1, \dots , \underline r_n)] & \to E[\underline{f(r_1,\dots,r_n)}] \\
  E[\tY \lambda x. M] & \to E[\lambda z. M[(\tY \lambda x. M)/x] z] \\
  & \qquad \qquad \text{ where $z$ is not free in $s$}\\
  E[\tif{\underline r < 0}{M_1}{M_2}] & \to E[M_1] \text{ where }r < 0 \\
  E[\tif{\underline r < 0}{M_1}{M_2}] & \to E[M_2] \text{ where }r \geq 0 \\
  E[\tsample] & \to E[\underline r] \text{ where } r \in [0,1].
\end{align*}
We write $\to^\ast$ for the reflexive, transitive closure of $\to$.
Every closed term either is a value or reduces to another term.

@@

Define a \emph{skeleton} to be a term but, instead of having real constants $\underline r$, it has a placeholder $\skeletonPlaceholder$, so that each term $M$ has a skeleton $\mathit{Sk}(M)$, and each skeleton $S$ can be converted to a term $S[r]$ given a vector $r$ of $n$ real numbers to substitute in, where $n$ is the number of occurrences of $\skeletonPlaceholder$ in $S$. 
Following \cite{DBLP:conf/icfp/BorgstromLGS16,DBLP:conf/lics/StatonYWHK16,DBLP:journals/pacmpl/EhrhardPT18}, the $\sigma$-algebra on $\Lambda$ is defined by identifying $\Lambda$ with $\bigcup_{m\geq 0} \big(\mathsf{Sk}_m \times \Real^m \big)$, where $\mathsf{Sk}_m$ is the set of skeletons containing $m$ occurrences of $\skeletonPlaceholder$.
Thus identified, we give $\Lambda$ the countable disjoint union topology of the product topology of the discrete topology on $\mathsf{Sk}_m$ and the standard topology on $\Real^m$, and take the corresponding Borel $\sigma$-algebra.
Note that the connected components of $\Lambda$ have the form $\{S\} \times \Real^m$, with $S$ ranging over $\mathsf{Sk}_m$, and $m$ over $\mathbb{N}$.

@@

%\paragraph{}
\begin{remark}
\label{rem:score}
Our definition of AST is equivalent to that given in \cite{MakOPW21} (although the program semantics is stated in a slightly different way), except for the presence of a $\tscore$ construct for soft conditioning. 
The $\tscore$ construct is irrelevant to termination except that it fails if its argument is negative, thus allowing computations to fail after finitely many steps. 
If such a thing were added, it would merely require an additional side condition to our main theorems, that no failing term was reachable (or, in the case of the confluent semantics based results, no term where the reduction strategy is not defined is reachable via the reduction strategy unless it's a value).
\end{remark}

### 2021-01-16 LICS21/antitone.tex
Introduction:ongoing Reformatted last antitone sparse ranking function in appendix

Define an \emph{antitone sparse ranking function} on a closed term $M$ to be a partial function $F : Rch(M) \rightharpoonup \mathbb{R}$ such that for some antitone function $\epsilon : \mathbb{R}_{\geq 0} \to \mathbb{R}_{>0}$:
\begin{inparaenum}[(i)]
\item $f(N) \geq 0$ for all $N \in \dom(f)$,
\item $M \in \dom(f)$,
\item for any $N \in \dom(f)$, evaluation of $N$ will eventually reach some $O$ which is either a value or in $\dom(f)$, and $f(N) \geq \mathbb E [f(O) + \epsilon(f(O)) \times \text{the number of $\tY$-reduction steps from $N$ to $O$}]$ (where $f(O)$ is taken to be 0 if $O$ is a value outside of $\dom(f)$).
\end{inparaenum}
\iffalse
\begin{itemize}
\item $f(N) \geq 0$ for all $N$ where $f$ is defined.
\item $f$ is defined at $M$.
\item For any $N$ in the domain of definition of $f$, evaluation of $N$ will eventually reach some $O$ which is either a value or in the domain of definition of $f$, and $f(N) \geq \mathbb E [f(O) + \epsilon(f(O)) \times \text{the number of $\tY$-reduction steps from $N$ to $O$}]$ (where $f(O)$ is taken to be 0 if $O$ is a value outside of the domain of $f$).
\end{itemize}
\fi
\begin{restatable}{theorem}{AntitonePartial}  \label{thm:antitone partial implies rankable}
  Every antitone sparse ranking function is a restriction of an antitone ranking function.
\end{restatable}

>>>>

Define an \emph{antitone sparse ranking function} on a closed term $M$ to be a partial function $F : Rch(M) \rightharpoonup \mathbb{R}$ such that for some antitone function $\epsilon : \mathbb{R}_{\geq 0} \to \mathbb{R}_{>0}$:
\begin{inparaenum}[(i)]
\item $f(N) \geq 0$ for all $N \in \dom(f)$,
\item $M \in \dom(f)$,
\item for any $N \in \dom(f)$, evaluation of $N$ will eventually reach some $O$ which is either a value or in $\dom(f)$, and $f(N) \geq \mathbb E [f(O) + \epsilon(f(O)) \times \text{the number of $\tY$-reduction steps from $N$ to $O$}]$ (where $f(O)$ is taken to be 0 if $O$ is a value outside of $\dom(f)$).
\end{inparaenum}
\iffalse
\begin{itemize}
\item $f(N) \geq 0$ for all $N$ where $f$ is defined.
\item $f$ is defined at $M$.
\item For any $N$ in the domain of definition of $f$, evaluation of $N$ will eventually reach some $O$ which is either a value or in the domain of definition of $f$, and $f(N) \geq \mathbb E [f(O) + \epsilon(f(O)) \times \text{the number of $\tY$-reduction steps from $N$ to $O$}]$ (where $f(O)$ is taken to be 0 if $O$ is a value outside of the domain of $f$).
\end{itemize}
\fi
\begin{restatable}[Sparse function]{theorem}{AntitonePartial}  \label{thm:antitone partial implies rankable}
  Every antitone sparse ranking function is a restriction of an antitone ranking function.
\end{restatable}

### 2021-01-16 LICS21/confluent.tex
Introduction:ongoing Reformatted last antitone sparse ranking function in appendix

\begin{example}[Labelling samples by terms]
Consider the terms
\begin{align*}
A[x] &= \tif{\tif{x>0}{I}{I} \, \tsample - 0.5 > 0}{0}{\Omega}\\
B &= \tif{\tsample - 0.5 > 0}{0}{\Omega}
\end{align*}
If terms rather than skeletons were used to label samples, the set of modified traces where $A[\tsample]$ terminates would be
\begin{align*}
\bigcup_{r \in [0,1]} \Big\{s \mid\ & s(A[\tsample], \mathsf{if}_1;\underline{-}_1;@_1;\mathsf{if}_1) = r, \\
& s(A[\tsample] \to A[\underline r] \to^* B, \mathsf{if}_1 ; \underline{-}_1) > 0.5 \Big\}.
\end{align*}

\iffalse
\lo{The occurrence of $L_s(M)$ in the set comprehension below should be $L_0(M)$?}

@@

\begin{definition}
The relation $\sim$ is defined as the union of the minimal symmetric relations $\sim_p$ (``$p$'' for parent-child) and $\sim_c$ (``$c$'' for cousin) satisfying
\begin{enumerate}
    \item If $N$ reduces to $O$ with the redex at position $\alpha$, and $\beta$ is a position in $N$ disjoint from $\alpha$, then $(N,\beta) \sim_p (O,\beta)$.

@@

        \item $N \mid \alpha = \tY (\lambda x. A)$, $A$ reduced at $\beta$ is $A'$, $(\gamma_i)_i$ is a list of all the positions where $A' \mid \gamma  = x$, ordered from left to right, $O_1$ is $N$ reduced first at $\alpha;\tY;\lambda;\beta$ then at $\alpha$, and $O_2$ is $N$ reduced first at $\alpha$ then at $\alpha;\lambda;@_1;\gamma_i;Y;\lambda;\beta$ for each $i$ in order where $\gamma_i$ is left of $\beta$ then at $\alpha;\lambda;@_1;\beta$ then at $\alpha;\lambda;@_1;\gamma_i;Y;\lambda;\beta$ for the remaining values of $i$.
    \end{enumerate}
    (in which case $O_1$ and $O_2$ are equal as skeletons, but with different reduction sequences), $O_1'$ and $O_2'$ are the results of applying some reduction sequence to each of $O_1$ and $O_2$ (the same reductions in each case, which is always possible because they're equal terms), and $\delta$ is a position in $O_1'$ (or equivalently $O_2'$), then $(O_1',\delta) \sim_c (O_2',\delta)$.
\end{enumerate}
The $\sim_c$-rules are illustrated in \Cref{sim-diagram}.
\end{definition}

@@

\begin{theorem} \label{thm:confluent ranking}
If a closed PPCF term $M$ has a ranking function, sparse ranking function, antitone ranking function or antitone sparse ranking function with respect to a reduction strategy $r$, then $M$ is AST with respect to $r$, and AST.
\end{theorem}

@@

\begin{restatable}{theorem}{confluentNoStronger}
For any closed term $M$ and reduction strategy $r$ on $M$, if every term in $\mathit{Rch}_r(M)$ is AST, then every term in $\mathit{Rch}(M)$ is AST.
\end{restatable}

>>>>

\begin{example}[Labelling samples by terms]
Consider the terms
\begin{align*}
A[x] &= \tif{\tif{x>0}{I}{I} \, \tsample - 0.5 > 0}{0}{\Omega}\\
B &= \tif{\tsample - 0.5 > 0}{0}{\Omega}
\end{align*}
If terms rather than skeletons were used to label samples, the set of modified traces where $A[\tsample]$ terminates would be
\[
\begin{array}{l}
\bigcup_{r \in [0,1]} \big\{s \mid\ s(A[\tsample], \mathsf{if}_1;\underline{-}_1;@_1;\mathsf{if}_1) = r, \\
\qquad\qquad\qquad s(A[\tsample] \to A[\underline r] \to^* B, \mathsf{if}_1 ; \underline{-}_1) > 0.5 \big\}.
\end{array}
\]
\iffalse
\lo{The occurrence of $L_s(M)$ in the set comprehension below should be $L_0(M)$?}

@@

\begin{definition}
The relation $\sim$ is defined as the union of the minimal symmetric relations $\sim_p$ (``$p$'' for parent-child) and $\sim_c$ (``$c$'' for cousin) satisfying
\begin{asparaenum}[(i)]
    \item If $N$ reduces to $O$ with the redex at position $\alpha$, and $\beta$ is a position in $N$ disjoint from $\alpha$, then $(N,\beta) \sim_p (O,\beta)$.

@@

        \item $N \mid \alpha = \tY (\lambda x. A)$, $A$ reduced at $\beta$ is $A'$, $(\gamma_i)_i$ is a list of all the positions where $A' \mid \gamma  = x$, ordered from left to right, $O_1$ is $N$ reduced first at $\alpha;\tY;\lambda;\beta$ then at $\alpha$, and $O_2$ is $N$ reduced first at $\alpha$ then at $\alpha;\lambda;@_1;\gamma_i;Y;\lambda;\beta$ for each $i$ in order where $\gamma_i$ is left of $\beta$ then at $\alpha;\lambda;@_1;\beta$ then at $\alpha;\lambda;@_1;\gamma_i;Y;\lambda;\beta$ for the remaining values of $i$.
    \end{enumerate}
    (in which case $O_1$ and $O_2$ are equal as skeletons, but with different reduction sequences), $O_1'$ and $O_2'$ are the results of applying some reduction sequence to each of $O_1$ and $O_2$ (the same reductions in each case, which is always possible because they're equal terms), and $\delta$ is a position in $O_1'$ (or equivalently $O_2'$), then $(O_1',\delta) \sim_c (O_2',\delta)$.
\end{asparaenum}
The $\sim_c$-rules are illustrated in \Cref{sim-diagram}.
\end{definition}

@@

\begin{theorem}[Confluent ranking] \label{thm:confluent ranking}
If a closed PPCF term $M$ has a ranking function, sparse ranking function, antitone ranking function or antitone sparse ranking function w.r.t.~a reduction strategy $r$, then $M$ is AST w.r.t.~$r$, and AST.
\end{theorem}

@@

\begin{restatable}{theorem}{confluentNoStronger}
\label{thm:confluentNoStronger}
For any closed term $M$ and reduction strategy $r$ on $M$, if every term in $\mathit{Rch}_r(M)$ is AST, then every term in $\mathit{Rch}(M)$ is AST.
\end{restatable}

### 2021-01-16 LICS21/intro.tex
Introduction:ongoing Reformatted last antitone sparse ranking function in appendix

@@

The language PPCF used in this paper is simply typed, with sampling from a continuous distribution, and an explicit recursion primitive, $\tY$. 
As the type system already constrains the terms enough to force termination in the absence of the recursion construct \cite{tait1967,BarendregtDS2010}, it is only the $\tY$-reduction steps that must be counted, making defining ranking functions somewhat easier.

We introduce the notion of \emph{sparse ranking function} to make constructing ranking functions more convenient. Most of the individual execution steps of a typical program are trivial and easy to mentally skip over. Sparse ranking functions can be defined only for those points in the execution of a program which are semantically important, while all of the other intermediate steps can be ignored. This also makes the ranking function method of proving AST more compatible with syntactic sugar, because the intermediate reduction steps implicit in the simplified notation can be ignored.

A ranking function (or sparse ranking function) provides a bound on the expected number of $\tY$-reduction steps before the program terminates, therefore ranking functions cannot be constructed for terms whose expected number of $\tY$-reduction steps is infinite, such as the simplest implementation of the 1D unbiased random walk. 
This restriction can be removed by generalising ranking functions to \emph{antitone ranking functions}, which rather than having to decrease by a constant amount for each $\tY$-reduction step, may decrease by a variable amount, depending on the value of the ranking function. 
This method is capable of proving the almost sure termination of programs which terminate arbitrarily slowly. Our progress condition is equivalent to that given by McIver et al.~\cite{DBLP:journals/pacmpl/McIverMKK18} (by applying a transformation to the ranking function), but needs to be different in order to be compatible with the presence of additional intermediate reduction steps where the ranking function is not required to decrease at all.

Basic (non-random) lambda calculus has the very useful Church-Rosser property, which implies (among other things) that even if execution of a program starts in a different order, it will still reach the same normal form eventually (assuming it does reach a normal form). 
Probabilistic lambda calculus does not have this property, because random choices may be duplicated, and evaluating the same subterm multiple times can yield different results. 
However, with a restricted set of reduction strategies, Church-Rosserness may be regained. 
We introduce a novel addressing scheme for the possible random choices in a program's execution, which ensures that the same random choices are taken at corresponding positions in alternative reduction sequences, so that the same eventual result can be reached. 
This is then used to prove yet another extension to the ranking function theorem, that ranking functions may be defined with respect to alternative reduction strategies (which in some cases may lead to a considerably simpler execution and ranking function), and rankability in this sense still imples almost sure termination. 
This confluent trace semantics has other possible applications as well, for example in Bayesian inference algorithms.

>>>>

\medskip

@@

The language PPCF used in this paper is simply typed, with random continuous sampling, and an explicit recursion primitive, $\tY$. 
We define a \emph{ranking function} on a term $M$ to be a non-negative measurable function on the reachable terms from $M$ whose expected value decreases or is unchanged as the term is reduced.
Because the type system already constrains the terms enough to force termination in the absence of the recursion construct \cite{tait1967,BarendregtDS2010}, it is only the $\tY$-reduction steps that must be counted, making defining ranking functions somewhat easier.
Using supermartingales and Doob's (iterated) Optional Sampling Theorem, we show the soundness of rankability (\cref{thm:rankable implies termination}): if a PPCF term has a ranking function then it is AST.

To define a ranking function, one has to try to organise the set of reachable terms (of the program being analysed) into a manageable number of syntactic cases. 
Unfortunately (for any fixed reduction strategy) there are programs whose reachable terms span such a large number of cases that it would be extremely difficult to analyse by hand, if at all possible. 
How can the construction of ranking functions be made possible or easier?

Our first answer is the notion of \emph{sparse ranking function}. 
Most of the individual execution steps of a typical program are trivial and easy to mentally skip over. Sparse ranking functions can be defined only for those points in the execution of a program which are semantically important, while all of the other intermediate steps can be ignored. Yet, they are no less efficacious for proving AST (\cref{thm:partial implies rankable}): every sparse ranking function is a restriction of a ranking function. This also makes the ranking function method of proving AST more compatible with syntactic sugar, because the intermediate reduction steps implicit in the simplified notation can be ignored.

A ranking function (or sparse ranking function) provides a bound on the expected number of $\tY$-reduction steps before the program terminates, therefore ranking functions cannot be constructed for terms whose expected number of $\tY$-reduction steps is infinite, such as the simplest implementation of the 1D unbiased random walk. 
This restriction can be removed by generalising ranking functions to \emph{antitone ranking functions}, which rather than having to decrease by a constant amount for each $\tY$-reduction step, may decrease by a variable amount, depending on the value of the ranking function. 
Thanks to this feature, the antitone ranking function method is capable of handling programs which terminate arbitrarily slowly. \lo{TODO: Reference some examples.}
We show that this method is sound for proving AST (\cref{thm:antitone rankable implies termination}), 
and it also enjoys a sparse function theorem (\cref{thm:antitone partial implies rankable}).
\akr{Our progress condition is equivalent to that given by McIver et al.~\cite{DBLP:journals/pacmpl/McIverMKK18} (by applying a transformation to the ranking function), but needs to be different in order to be compatible with the presence of additional intermediate reduction steps where the ranking function is not required to decrease at all.}
\lo{The same point is made in \cref{sec:related}, so the preceding can be omitted.}

Basic (non-random) lambda calculus has the very useful Church-Rosser property, which implies (among other things) that even if execution of a program starts in a different order, it will still reach the same normal form eventually (assuming it does reach a normal form). 
Probabilistic lambda calculus does not have this property, because random choices may be duplicated, and evaluating the same subterm multiple times can yield different results. 
However, with a restricted set of reduction strategies, Church-Rosserness may be regained. 
We introduce a novel addressing scheme for the possible random choices in a program's execution, which ensures that the same random choices are taken at corresponding positions in alternative reduction sequences, so that the same eventual result can be reached. 
This is then used to prove yet another extension to the ranking function theorem, that (\cref{thm:confluent ranking}) ranking functions may be defined with respect to alternative reduction strategies (which in some cases may lead to a considerably simpler execution and ranking function), and (\cref{thm:confluentNoStronger}) rankability in this sense still imples almost sure termination. 
We stress that the confluent trace semantics is of wider interest, and has other possible applications as well, for example in Bayesian inference algorithms.

By combining these methods we can prove AST of a variety of PPCF programs, including cases (e.g.~\cref{ex:non-affine recursion,ex:higher-order recursion}) that are beyond the reach of methods in the literature.

### 2021-01-16 LICS21/ranking.tex
Introduction:ongoing Reformatted last antitone sparse ranking function in appendix

\begin{restatable}{theorem}{PartialImpliesRankable} \label{thm:partial implies rankable}
Every sparse ranking function is a restriction of a ranking function.
\end{restatable}

>>>>

\begin{restatable}[Sparse function]{theorem}{PartialImpliesRankable} \label{thm:partial implies rankable}
Every sparse ranking function is a restriction of a ranking function.
\end{restatable}

### 2021-01-16 LICS21/supermartingales.tex
Introduction:ongoing Reformatted last antitone sparse ranking function in appendix

\begin{remark}
Our notion of ranking supermartingale is a slight generalisation of the original definition, due to \cite{DBLP:conf/cav/ChakarovS13,DBLP:conf/popl/FioritiH15}, which does not refer to a stopping time explicitly:
Fioriti and Hermanns call a supermartingale $(Y_n)_{n \geq 0}$ \emph{ranking} if $Y_n \geq 0$, and there exists $\epsilon > 0$ such that $\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon \cdot {\bf 1}_{\set{Y_n > 0}}$.
\end{remark}

@@

In the rest of this section we show that if $M \in \Lambda^0$ is rankable, then $M$ is AST: in other words, the method of ranking function is sound for proving a.s.~termination of PPCF programs.

Let $f$ be a ranking function on $M$.
Define random variables on the probability space $(\entrosp, \Sigma_\entrosp, \mu_{\entrosp})$:
\begin{align}
T_{-1}(s) & := -1 \nonumber \\
T_{n+1}(s) & := \min \{ k \mid k>T_n(s), M_k(s) \textrm{ a value} \nonumber\\
& \qquad\qquad\qquad\qquad\qquad \hbox{or of form } E[\tY \lambda x. N] \} \nonumber \\
Y_n(s) & := f(M_{T_n}(s)) \nonumber \\
T_M^\tY(s) &:= \min \set{n \mid M_{T_n(s)}(s) \textrm{ is a value}} \label{eq:y stopping time} 
\end{align}

@@

The random variable $T_M^\tY$, which we call the \emph{$\tY$-runtime of $M$}, can equivalently be defined as the number of $\tY$-reduction steps in the reduction sequence of $M$. 
Note that, as the type system ensures that the reduction relation excluding $\tY$-reduction is strongly normalising, only finitely many reductions can occur in a row without one of them being a $\tY$-reduction, therefore $T_M^\tY < \infty$ a.s.~iff $M$ is AST. 
Moreover 
\begin{restatable}{lemma}{TMtYStoppingTime}
\label{lem:TMtY is a stopping time}
$T_M^\tY$ is a stopping time adapted to $(\calF_{T_n})_{n \geq 0}$.
\end{restatable}

@@

\begin{proof}
Let $f$ be a ranking function on $M$.
For 1, since $(T_n)_{n \geq 0}$ is an increasing sequence of stopping times, each is adapted to $(\calF_n)_{n \geq 0}$ and bounded (\Cref{lem:TnBounded}),
and $(f(M_n))_{n \geq 0}$ is a supermartingale also adapted to $(\calF_n)_{n \geq 0}$ (\Cref{thm:rankable and strict rankable}),
it follows from the Optional Sampling \Cref{thm:optional sampling} that $(Y_n)_{n \geq 0}$ %where $Y_n := f(M_{T_n})$ 
is a supermartingale adapted to $(\calF_{T_n})_{n \geq 0}$.
Notice the stopping time $T_M^\tY$ is also adapted to $(\calF_{T_n})_{n \geq 0}$ (\Cref{lem:TMtY is a stopping time}); and we have that $(Y_n)_{n \geq 0}$ is a $1$-ranking supermartingale.
Therefore, by \Cref{lem:rank-PAST}, $T^{\tY}_M < \infty$ a.s.~and $\expect{T^{\tY}_M} < \infty$.

%For 2, we have that $(f(M_n))_{n \geq 0}$ is a ranking supermartingale (\Cref{thm:rankable and strict rankable}) w.r.t.~stopping time $T_M$. 
%It follows immediately from \Cref{lem:rank-PAST} that $\expect{T_M} < \infty$.
Statement 2 follows from \Cref{thm:rankable and strict rankable} and \Cref{lem:rank-PAST}.
\end{proof}

>>>>

\begin{remark}
Our notion of ranking supermartingale is a minor generalisation of the original definition, due to \cite{DBLP:conf/cav/ChakarovS13,DBLP:conf/popl/FioritiH15}, which does not refer to a stopping time explicitly.
%: Fioriti and Hermanns call a supermartingale $(Y_n)_{n \geq 0}$ \emph{ranking} if $Y_n \geq 0$, and there exists $\epsilon > 0$ such that $\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon \cdot {\bf 1}_{\set{Y_n > 0}}$.
\end{remark}

@@

In the rest of this section we show that if a PPCF term is rankable, then it is AST. 
%in other words, the method of ranking function is sound for proving a.s.~termination of PPCF programs.

Let $f$ be a ranking function on $M \in \Lambda^0$.
Define random variables on the probability space $(\entrosp, \Sigma_\entrosp, \mu_{\entrosp})$:
\begin{align}
T_{-1}(s) & := -1 \nonumber \\
T_{n+1}(s) & := \min \{ k \mid k>T_n(s), M_k(s) \textrm{ a value} \nonumber\\
& \qquad\qquad\qquad\qquad\qquad \hbox{or of form } E[\tY \lambda x. N] \} \nonumber \\
Y_n(s) & := f(M_{T_n}(s)) \nonumber \\
T_M^\tY(s) &:= \min \set{n \mid M_{T_n(s)}(s) \textrm{ is a value}} \label{eq:y stopping time} 
\end{align}

@@

The random variable $T_M^\tY$, which we call the \emph{$\tY$-runtime of $M$}, can equivalently be defined as the number of $\tY$-reduction steps in the reduction sequence of $M$. 
Note that, as the type system ensures that the reduction relation excluding $\tY$-reduction is strongly normalising, only finitely many reductions can occur in a row without one of them being a $\tY$-reduction, therefore $T_M^\tY < \infty$ a.s.~iff $M$ is AST. 
Moreover: 
\begin{restatable}{lemma}{TMtYStoppingTime}
\label{lem:TMtY is a stopping time}
$T_M^\tY$ is a stopping time adapted to $(\calF_{T_n})_{n \geq 0}$.
\end{restatable}

@@

\begin{proof}
Let $f$ be a ranking function on $M$.
For (i), since $(T_n)_{n \geq 0}$ is an increasing sequence of stopping times, each is adapted to $(\calF_n)_{n \geq 0}$ and bounded (\Cref{lem:TnBounded}),
and $(f(M_n))_{n \geq 0}$ is a supermartingale also adapted to $(\calF_n)_{n \geq 0}$ (\Cref{thm:rankable and strict rankable}),
it follows from the Optional Sampling \Cref{thm:optional sampling} that $(Y_n)_{n \geq 0}$ %where $Y_n := f(M_{T_n})$ 
is a supermartingale adapted to $(\calF_{T_n})_{n \geq 0}$.
Notice the stopping time $T_M^\tY$ is also adapted to $(\calF_{T_n})_{n \geq 0}$ (\Cref{lem:TMtY is a stopping time}); and we have that $(Y_n)_{n \geq 0}$ is a $1$-ranking supermartingale.
Therefore, by \Cref{lem:rank-PAST}, $T^{\tY}_M < \infty$ a.s.~and $\expect{T^{\tY}_M} < \infty$.
%For 2, we have that $(f(M_n))_{n \geq 0}$ is a ranking supermartingale (\Cref{thm:rankable and strict rankable}) w.r.t.~stopping time $T_M$. 
%It follows immediately from \Cref{lem:rank-PAST} that $\expect{T_M} < \infty$.
Statement (ii) follows at once from \Cref{thm:rankable and strict rankable} and \Cref{lem:rank-PAST}.
\end{proof}
